Dunnett's test

### 2022-09-06 text/22/analysis-of-means.Rmd
Updated some of the 2017 files

---
title: "A second alternative to the pairwise testing approach"
author: "Steve Simon"
date: '2022-07-04'
output:
  word_document: default
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

@@

You are in a setting where you are expecting to see a degree of uniformity among several groups. For example, you have a group of clinics who all treat the same type of patients. The patients themselves may differ markedly from one to another, but they are assigned in a more or less random way to each clinic. So, although one patient may differ markedly from another, the average result should be the same for across all the clinics.

The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient. But because of how the patients are assigned to a particular clinic, these should balance out on average. If there is one clinic that has different average results, it has to be because that the physicians at that clinic have a different approach to treatment. If you can identify a clinic (it could be more than one clinic, of course) that differs from the norm, then investigate. Perhaps that clinic is doing something "wrong" that causes the discrepancy, or maybe that clinic is the only one doing things right. In any case understanding why you see unexpected deviations in a process that should be uniform across clinics provides you with an opportunity to learn something.

@@

Suppose you are in a setting like the one described above, where there is an expectation of similarity between each group mean and the overall mean. Then you would write your hypothesis as 
```{r, fig.width=5, fig.height=0.33, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.5, h0, adj=0)
```

@@

The analysis of means approach that compares each group mean to the overall mean is easy to implement and it lends itself to a simple graphical display.

@@

It's important to editorialize a bit here. Deviating from the norm could be a good thing or a bad thing or it could be an indifferent thing. Your goal is not to use statistics to hunt out different groups to reward or punish them. You are using statistics to help in understanding why deviations from the norm occur.

@@

Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant. Four plants were assessed for each strain. Here are the descriptive statistics. The research question is whether all strains have a comparable amount of CryA1c in their leaves.

@@

@@

@@

>>>>

---
title: "A second alternative to the pairwise testing approach"
author: "Steve Simon"
date: '2022-07-04'
output:
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

@@

### Comparison to an overall mean

You are in a setting where you are expecting to see a degree of uniformity among several groups. For example, you have a group of clinics who all treat the same type of patients. The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient. But because of how the patients are assigned to a particular clinic (more or less randomly), these should balance out on average.

If there is one clinic that has different average results, it has to be because that the physicians at that clinic have a different approach to treatment. If you can identify a clinic (it could be more than one clinic, of course) that differs from the norm, then investigate. Perhaps that clinic is doing something "wrong" that causes the discrepancy, or maybe that clinic is the only one doing things right. In any case understanding why you see unexpected deviations in a process that should be uniform across clinics provides you with an opportunity to learn something.

@@

Suppose you are in a setting like the one described above, where there is an expectation of similarity between each group mean and the overall mean. Then you would write your hypothesis as

```{r, fig.width=5, fig.height=0.33, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.5, h0, adj=0)
```

@@

The analysis of means approach compares each group mean to the overall mean. It is easy to implement and it lends itself to a simple graphical display. You need a table of critical values, which depend on alpha (the overall Type I error rate), t (the number of groups), and n (the number of observations within each group). Some tables use the degrees of freedom for error in place of n.

@@

It's important to editorialize a bit here. Deviating from the norm could be a good thing or a bad thing or it could be an indifferent thing. Your goal is not to use statistics to hunt out different groups to reward or punish them. You are using statistics to help in understanding if deviations from the norm occur and then study those deviating groups to understand why they deviate.

@@

Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant. Four plants were assessed for each strain. Here are the descriptive statistics. The research question is whether any strain differs from the overall mean level of CryA1c.

@@

You may prefer a one-sided test in this setting, such as testing whether any strain is deficient in the CryA1c levels. You can make a very easy modification to get one-sided tests.

@@

You also need to specify the Analysis of Means hypothesis prior to looking at your data. Peeking at the data and then choosing your hypothesis is cheating.

@@

### Details about the calculations

To get the limits for the graph shown above, you need the Analysis of Means table of critical values plus an estimate of Mean Squared Error. The general formula (as shown in [this article][jay1]) is

$\ $ $\bar Y_{..} \pm h(\alpha, t, N) \sqrt{(t-1)MSE/N}$

where $\bar Y_{..}$ is the overall mean and $h(\alpha, t, N)$ comes from a table such as [this one][lev1] (see page 3) or [this one][sim3].

In our example alpha=0.05, t=5, and N=20. The original article does not provide MSE, but you can calculate it as the pooled variance using the descriptive statistics shown above. The limits are

```{r, echo=FALSE}
# Make the numbers look nice
ybarbar <- round(ybarbar, 2)
mse <- round(mse, 2)
lo <- round(lo, 2)
hi <- round(hi, 2)
```

$\ $ $`r ybarbar` \pm 2.88 \sqrt{(`r t`-1)`r mse`/`r N`}$

or

$\ $ $`r lo`$ and $`r hi`$

[jay1]: https://digitalcommons.pvamu.edu/cgi/viewcontent.cgi?article=1122&context=aam

[lev1]: https://wps.prenhall.com/wps/media/objects/9431/9657451/Ch_11/levine-smume6_topic_ANOM.pdf

[sim3]: http://www.pmean.com/07/AnomTable05Part1.html

### 2022-09-06 text/22/dunnetts-test.Rmd
Updated some of the 2017 files

---
title: "Dunnett's test"
author: "Steve Simon"
date: '2022-07-04'
output:
  word_document: default
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

@@

Dunnett's test is easy to implement. Calculate the traditional measures in analysis of variance, including mean squared error (MSE). You can skip the traditional F Test that you see in most ANOVA tables. It won't hurt your overall alpha level, as long as you only look at each treatment versus the control. Simply compare each treatment mean minus the control mean to a cutoff value using percentiles from a special table for Dunnett's test. The critical value for Dunnett's test depends on alpha (the desired Type I error rate for a two-sided test), g (the number of groups including the control group) and n (the number of data points in each group). Note that some tables define g as the number of groups excluding the control group.

Most statistical software packages will include Dunnett's test as an option for ANOVA. Just remember that you don't need to use the initial F test as a screen before jumping into the Dunnett's test.

@@

@@

Disinfection is important, but does it reduce the tensile strength of the ligatures. Here are the summary statistics for the maximum strength measurement.

@@

The paper does not include the raw data, but you can still calculate all the key statistics needed to run Dunnett's test. You can create a simple graph that identifies if any of the disinfection protocols shows significantly less strength than the control.

@@

@@

Let's consider this as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.

>>>>

---
title: "Dunnett's test"
author: "Steve Simon"
date: '2022-07-04'
output:
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

@@

Dunnett's test is easy to implement. Calculate the traditional measures in analysis of variance, including mean squared error (MSE). You can skip the F Test that you see in most ANOVA tables. It won't hurt your overall alpha level, as long as you only look at each treatment versus the control. Simply compare each treatment mean minus the control mean to a cutoff value using percentiles from a special table for Dunnett's test. The critical value for Dunnett's test depends on alpha (the desired Type I error rate for a two-sided test), g (the number of groups including the control group) and n (the number of data points in each group). Note that some tables define g as the number of groups excluding the control group.

Most statistical software packages will include Dunnett's test as an option for ANOVA. Just remember that you don't need to use the initial F test as a screen before jumping into Dunnett's test.

@@

(Image taken from G&H Orthodontics)

@@

Disinfection is important, but does it reduce the tensile strength of the ligatures? Here are the summary statistics for the maximum strength measurement.

@@

The paper does not include the raw data, but you can still calculate all the key statistics needed to run Dunnett's test. You can then create a simple graph that identifies if any of the disinfection protocols shows significantly less strength than the control.

@@

The dotted line is the mean for the control group. Any mean falling outside the solid lines is statistically significantly different from the control mean. Notice that there is no comparison at 1 because you don't compare the control group to itself.

@@

Let's consider this as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings. If, however, you are truly only interested in comparing each treatment to the control and you specify this prior to looking at the data, then Dunnett's test is a very attractive option.

### Details about the calculations

The limits require a critical value from a table for Dunnett's test, the mean of the control group and Mean Squared Error.

$\ $ $\bar X_1 \pm d(\alpha, t, n) \sqrt{2 MSE/n}$

where $\alpha$ is the Type I error rate, t in the number of groups, and n is the number of observations in each group. The value of d can be found in various tables, such as [this one][bob1] or [this one][win1]. While the paper that I drew the example from does not include the raw data, you can calculate Mean Squared Error from the summary statistics. The limits in this example are

```{r echo=FALSE}
# Make the numbers look nice.
mse <- round(mse, 2)
lo <- round(lo, 2)
hi <- round(hi, 2)
```

$\ $ $`r mn1[1]` \pm 2.73 \sqrt{2 (`r mse`)/`r n`}$

$\ $ $`r lo`$ and $`r hi`$

[bob1]: https://www.statology.org/dunnetts-table/
[win1]: https://users.stat.ufl.edu/~winner/tables/dunnett-2side.pdf
